Efficient toolkit implementing best practices for principal component analysis of population genetic data

### 2019-11-07 paper-bedpca.tex
save

\renewcommand{\arraystretch}{2}

@@

Principal Component Analysis (PCA) has been widely used in genetics has been used for many years and in many contexts such as e.g.\ to adjust for population structure in Genome-Wide Association Studies (GWAS) by adding PCs as covariates \cite[]{price2006principal} and to detect loci under selection based on PCA \cite[]{galinsky2016fast,luu2017pcadapt}.
Recently, the advent of large population-scale genetic datasets, such as the UK biobank data, has prompted research on developing scalable algorithms to compute PCA on data as large as the UK Biobank \cite[]{bycroft2017genome}.
This is now possible thanks to software such as FastPCA (fast mode), FlashPCA2, PLINK 2.0 (approx mode), bigstatsr/bigsnpr and TeraPCA \cite[]{galinsky2016fast,abraham2017flashpca2,chang2015second,prive2017efficient,bose2019terapca}.

However, some pitfalls related to PCA of genotype data have been documented and not all of the currently available software addresses all of these. 
In the following, we recall common pitfalls and explain when they are relevant.
First, efficient methods for PCA use approximations, which can results in some lack of precision of computed PCs. This has been shown to be a possible issue for software such as FastPCA and PLINK 2.0, but not for FlashPCA2 and bigstatsr/bigsnpr \cite[]{abraham2017flashpca2,prive2017efficient}.
Second, some of the PCs may capture LD structure rather than population structure \cite[]{price2008long,abdellaoui2013population,prive2017efficient}. Including PCs that capture LD as covariates in genetic analyses can lead to reduced power, e.g.\ for detecting genetic association within these LD regions in GWAS.  
Third, another issue may arise when projecting PCs of a reference dataset to another study dataset: projected PCs are shrunk towards 0 in the new dataset \cite[]{lee2010convergence,wang2015improved,zhang2019fast}. This shrinkage makes it dangerous to use the projected PCs for analyses such as PC regression, ancestry detection and correction for ancestry.
Finally, PC scores may capture outliers that are due to family structure, population structure or other reasons; it might be benificial to detect and remove these individuals to maximise the population structure captured by the PCA (in the case of removing a few outliers) or to restrict analyses to genetically homogeneous samples.

In this paper, we derive a new implementation of PCA that can be used directly on bed/bim/fam PLINK files with some missing values, which we make available in R package bigsnpr v1.0.0. Then, we recall some of the pitfalls of PCA computation on genotype data and provide efficient solutions to these issues. These includes accouting for LD structure in PCA, detecting outliers and projecting PCs to another study sample without shrinkage bias.

[SHOULD ALSO TALK ABOUT PCADAPT -> CONFOUNDED BY LD + SHOULD COMPUTE MAHA ON UNLINKED]

@@

\begin{table}[h]
\centering
\caption{Overview of existing methods. [TODO: ADD CITATIONS + RESOLVE QUESTIONS]} 
\label{tab:method-overview}
\begin{tabular}{|c|c|c|c|}
\hline
Analysis & Method / Software & Pros & Cons \\
\hline
\multirow{3}{*}{PCA} & bigstatsr / bigsnpr \cite[]{prive2017efficient} & \makecell{Fast and accurate + handle dosages \\ + thinning options directly included} & \makecell{Own format without missing values \\ (fast functions are available \\ for converting and imputing)} \\
 & FlashPCA2 \cite[]{abraham2017flashpca2} & Fast and accurate & Sequential implementation \\
 & \makecell{PLINK 2.0 (reimplementation of FastPCA) \\ \cite[]{galinsky2016fast,chang2015second}} & Very fast (AVX2 only?) & \makecell{Not always 100\% accurate \\ \cite[]{abraham2017flashpca2,prive2017efficient}} \\
\hline
\makecell{Detection of \\ oultier samples} & \makecell{"6 SDs from the mean" \\ (EIGENSOFT, \cite{patterson2006population})} & Simple & Assume Gaussian distritbution \\
\hline
\makecell{Detection of \\ homogeneous \\ samples} & Aberrant \cite[]{bellenguez2011robust} & Robust & Only two statistics at once? \\
\hline
\multirow{5}{*}{\makecell{Projection of \\ new individuals \\ onto reference \\ PCA space}} & Simple projection (multiplication by loadings) & Simple & Shrinkage biased \\
  & Bias-adjusted projection \cite[]{dey2019asymptotic} & Independent of new samples & \makecell{Assume same shrinkage for all individuals \\ + model-based \\ + need all eigenvalues of reference} \\
  & \makecell{Augmentation, Decomposition, \\ and Procrustes transformation \\ (ADP, \cite{wang2015improved})} & Accurate & Slow \\
  & Online ADP (OADP, \cite{zhang2019fast}) & Much faster than ADP & Does not work for related individuals \\ 
\hline
\end{tabular}
\end{table}

@@

@@

For detecting outlier variants in PCA that are due to long-range Linkage Disequilibrium (LD) regions, we use a slightly modified version of the procedure described by \cite{prive2017efficient}.
We first apply a first round of clumping, prioritising variants by higher minor allele count, at e.g.\ $r^2$ > 0.2.
Then, we compute $K$ PC scores and loadings.
To summarise the contribution of each variant in all $K$ PC loadings, we compute the robust Mahalanobis distances (Section \ref{maha}) of these PC loadings.  
This captures both population-differentiating variants as well as long-range LD regions; to capture consecutive outliers that corresponds to long-range LD regions, we apply a Gaussian smoothing to these statistics.

[FIGURE TO ILLUSTRATE THIS?]

@@

For detecting outlier samples in PCA, we use a modified version of the Probabilistic Local Outlier Factor (PLOF) statistic on PCs \cite[]{kriegel2009loop}. Using K nearest neighbours (KNN), this consists in comparing the distance from an point $j$ to its KNNs ($pd_j = \sqrt{\frac{1}{K} \sum_{k=1}^K d^2_{j \leftrightarrow j_k}}$, where $j_k$ is the $k$-th NN of $j$) with the distances from its KNNs to their KNNs ($\frac{1}{K} \sum_{k=1}^K pd_{j_k}$). The idea is that an outlier should be far from all other points, but this point is even more outlier if its KNNs are in a very dense cluster. 
In practice, we use $pd_j / \sqrt{\frac{1}{K} \sum_{k=1}^K pd_{j_k}}$ (note the squared root) as it otherwise detect any point next to a very dense cluster as an outlier.
We implement this statistic in function \texttt{prob\_dist} of our new R package bigutilsr. To make it fast, we use and parallelise (\texttt{knn\_parallel}) the fast K nearest neighbours implementation of R package nabor \cite[]{elseberg2012comparison}.
Automatic detection of outliers is very hard, then we recommend users to choose a threshold of outlierness for this statistic based on visual inspection.

For detecting samples that have a different ancestry from most of the samples in the data, we compute the pairwise orthogonalized Gnanadesikan-Kettenrin robust Mahalanobis distance on PC scores (Section \ref{maha}). We then restrict to individuals whose log-distance (alternatively p-value) is smaller (larger) than some threshold determined based on visual inspection.

@@

To project PCs of a reference dataset (e.g.\ 1000G) to a target genotype dataset, we implement the following 3 steps in function \texttt{bed\_projectPCA} of package bigsnpr: 1) matching the variants of each dataset, including removing ambiguous alleles [A/T] and [C/G], and matching strand and direction of the alleles; 2) computing PCA of the reference dataset using the matched variants only; 3) projecting computed PCs to the target data using an optimised implementation (see Supplementary Materials) of the Online Augmentation, Decomposition, and Procrustes (OADP) transformation \cite[]{zhang2019fast}.

@@

We provide a subsetted version of the 1000 genomes (1000G) project data \cite[]{10002015global,meyer2019genotype}.
Variants are restricted to the ones in common with HapMap3 and UK Biobank \cite[]{international2010integrating,bycroft2017genome}. 
Moreover, we apply some quality control filters; we remove variants having a minor allele frequency < 0.01, variants or individuals with more than 10\% missing value, variants with P-value of the Hardy-Weinberg exact test < $10^{-50}$, and non-autosomal variants. 
To remove related individuals with second-degree relationship or more, we apply KING cutoff of 0.0884 to the data using PLINK 2.0 \cite[]{manichaikul2010robust,chang2015second}.
This results in 2490 individuals of the 1000G project (phase 3) in PLINK bed/bim/fam format. 
Resulting PLINK files and R code to generate these files are available at \url{https://doi.org/10.6084/m9.figshare.9208979.v3}. 
To easily download these data, we provide a function called \texttt{download\_1000G} in R package bigsnpr.

[REMOVE QC NA FOR 1000G?]

@@

\noindent[TALK ABOUT REMOVING OF LRLDR?]

@@

To show that we provide very fast implementations of the different methods presented in this paper, we apply them to the UK Biobank. We use 20 physical cores for most of the computations.
It takes 22 minutes to perform a first phase of clumping on 406,545 unrelated individuals genotyped over 504,139 variants, which reduces the number of variants to 261,307. It then takes 34 minutes to compute the first 20 PCs using these 261,307 variants.
If performing the automatic procedure for LD detection, it takes 5 hours to perform the initial clumping step, 6 computation of PCs and 5 outlier variant detection (i.e.\ 5 iterations of outlier detection and one final computation of PCs).

@@

Finally, it takes only 6 minutes to compute the 30-Nearest Neighbours of PC scores for all UK Biobank individuals, which is the main computational aspect of computing statistics for detection individual outlier samples (Section \ref{outlier-sample}).

@@

Then, we computed again PCA after restricting the individuals included in the computations: we randomly subsampled UKBB data to use only 10,000 British individuals (out of 431,029) and 5000 Irish individuals (out of 12,755), while keeping all individuals with other self-reported individuals. 
We further removed all pairs of related individuals reported by the UKBB (both individuals in each pair). 
This resulted in 48,942 individuals that we used to compute 50 PCs, which took less than 3 hours using function \texttt{bed\_autoSVD} (that converged after 4 iterations of automatic LD removal). 
We show that we are able to capture at least 40 PCs that display visual population structure (Figures \ref{fig:UKBB-scores2} and \ref{fig:UKBB-screeplot2}).
We then projected all 439,429 remaining individuals from UKBB onto this PCA space in 41 minutes. Note that these individuals are not supposed to be related to any of the 48,942 individuals used for training PCA because we removed both individuals from each pair of related individuals in UKBB. Projection of new individuals show again clear shrinkage when using simple projection (between 1.00 for PC1 and 1.80 for PC50), but no visible bias when using OADP projection (Figure \ref{fig:projUKBB-related}).

@@

\noindent[SHOW THAT PROCEDURE IS OKAY FOR SMALLER DATASETS?]

\noindent[SHOW ABLE TO RECONSTRUCT PCA? MAYBE NOT PC17]

@@

@@

\begin{figure}[!htpb]
\centerline{\includegraphics[width=0.8\textwidth]{UKBB-screeplot.pdf}}
\caption{Scree plot: plot of singular values computed from the UK Biobank.
\label{fig:UKBB-screeplot}}
\end{figure}

\begin{figure}[!htpb]
\centerline{\includegraphics[width=0.9\textwidth]{UKBB-PC1-20.png}}
\caption{Principal Component (PC) scores 1 to 20 computed from the UK Biobank.
\label{fig:UKBB-scores}}
\end{figure}

\begin{figure}[!htpb]
\centerline{\includegraphics[width=0.9\textwidth]{UKBB-loadings.pdf}}
\caption{Principal Component (PC) loadings 1 to 20 computed from the UK Biobank.
\label{fig:UKBB-loadings}}
\end{figure}

@@

\begin{figure}[!htpb]
\centerline{\includegraphics[width=0.8\textwidth]{UKBB-screeplot-restricted.pdf}}
\caption{Scree plot: plot of singular values computed from the UK Biobank using 48,942 individuals of diverse ancestries.
\label{fig:UKBB-screeplot2}}
\end{figure}

\begin{figure}[!htpb]
\centerline{\includegraphics[width=0.9\textwidth]{UKBB-scores-restricted.png}}
\caption{Principal Component (PC) scores 27 to 50 computed from the UK Biobank using 48,942 individuals of diverse ancestries. Different colors represent different self-reported ancestries.
\label{fig:UKBB-scores2}}
\end{figure}

>>>>

%\renewcommand{\arraystretch}{1.8}

@@

Principal Component Analysis (PCA) has been widely used in genetics for many years and in many contexts. For instance, adding PCs as covariates has been used to adjust for population structure in Genome-Wide Association Studies (GWAS) \cite[]{price2006principal}. PCA has also been used to detect loci under selection \cite[]{galinsky2016fast,luu2017pcadapt}.
Recently, the advent of large population-scale genetic datasets, such as the UK biobank data, has prompted research on developing scalable algorithms to compute PCA on data as large as the UK Biobank \cite[]{bycroft2017genome}.
This is now possible thanks to software such as FastPCA (fast mode of EIGENSOFT), FlashPCA2, PLINK 2.0 (approx mode), bigstatsr/bigsnpr and TeraPCA \cite[]{galinsky2016fast,abraham2017flashpca2,chang2015second,prive2017efficient,bose2019terapca}.

However, some pitfalls related to PCA of genotype data have been documented and not all of the currently available software address all of these. 
In the following, we recall common pitfalls and explain when they are relevant.
First, efficient methods for PCA use approximations, which can results in some lack of precision of computed PCs. This has been shown to be a possible issue for software such as FastPCA and PLINK 2.0, but not for FlashPCA2 and bigstatsr/bigsnpr \cite[]{abraham2017flashpca2,prive2017efficient}.
Second, some of the PCs may capture LD structure rather than population structure \cite[]{price2008long,abdellaoui2013population,prive2017efficient}. Including PCs that capture LD as covariates in genetic analyses can lead to reduced power, e.g.\ for detecting genetic association within these LD regions in GWAS.  
Third, another issue may arise when projecting PCs of a reference dataset to another study dataset: projected PCs are shrunk towards 0 in the new dataset \cite[]{lee2010convergence,wang2015improved,zhang2019fast}. This shrinkage makes it potentially dangerous to use the projected PCs for analyses such as PC regression, ancestry detection and correction for ancestry.
Finally, PC scores may capture outliers that are due to family structure, population structure or other reasons; it might be benificial to detect and remove these individuals to maximise the population structure captured by the PCA (in the case of removing a few outliers) or to restrict analyses to genetically homogeneous samples.

In this paper, we derive a new implementation of PCA that can be used directly on bed/bim/fam PLINK files with some missing values, which we make available in a new release of R package bigsnpr (v1.0.0). Then, we recall some of the pitfalls of PCA computation on genotype data and provide efficient solutions to these issues, which we implement in R packages bigsnpr and bigutilsr. These includes accouting for LD structure in PCA, detecting outliers and projecting PCs to another study sample without shrinkage bias.

@@

\begingroup
\renewcommand*{\arraystretch}{2}
\begin{table}[h]
\centering
\caption{Overview of existing methods. [TODO: ADD CITATIONS + RESOLVE QUESTIONS]} 
\label{tab:method-overview}
\begin{tabular}{|c|c|c|c|}
\hline
Analysis & Method / Software & Pros & Cons \\
\hline
\multirow{3}{*}{PCA} & bigstatsr / bigsnpr \cite[]{prive2017efficient} & \makecell{Fast and accurate + handle dosages \\ + thinning options directly included} & \makecell{Own format without missing values \\ (fast functions are available \\ for converting and imputing)} \\
 & FlashPCA2 \cite[]{abraham2017flashpca2} & Fast and accurate & Sequential implementation \\
 & \makecell{PLINK 2.0 (reimplementation of FastPCA) \\ \cite[]{galinsky2016fast,chang2015second}} & Very fast (AVX2 only?) & \makecell{Not always 100\% accurate \\ \cite[]{abraham2017flashpca2,prive2017efficient}} \\
\hline
\makecell{Detection of \\ oultier samples} & \makecell{"6 SDs from the mean" \\ (EIGENSOFT, \cite{patterson2006population})} & Simple & Assume Gaussian distritbution \\
\hline
\makecell{Detection of \\ homogeneous \\ samples} & Aberrant \cite[]{bellenguez2011robust} & Robust & Only two statistics at once \\
\hline
\multirow{5}{*}{\makecell{Projection of \\ new individuals \\ onto reference \\ PCA space}} & Simple projection (multiplication by loadings) & Simple & Shrinkage biased \\
  & Bias-adjusted projection \cite[]{dey2019asymptotic} & Independent of new samples & \makecell{Assume same shrinkage for all individuals \\ + model-based \\ + need all eigenvalues of reference} \\
  & \makecell{Augmentation, Decomposition, \\ and Procrustes transformation \\ (ADP, \cite{wang2015improved})} & Accurate & Slow \\
  & Online ADP (OADP, \cite{zhang2019fast}) & Much faster than ADP & Does not work for related individuals \\ 
\hline
\end{tabular}
\end{table}
\endgroup

@@

When computing these distances on $K$ approximately independent variables such as the first $K$ PC scores, the computed squared distances follow a $\chi^2(K)$ distribution.

@@

For detecting outlier variants in PCA that are due to long-range Linkage Disequilibrium (LD) regions, we use a similar procedure as described by \cite{prive2017efficient}.
We first apply a first round of clumping, prioritising variants by higher minor allele count, at e.g.\ $r^2$ > 0.2.
Then, we compute $K$ PC scores and loadings.
To summarise the contribution of each variant in all $K$ PC loadings, we compute the robust Mahalanobis distances (Section \ref{maha}) of these PC loadings.  
This captures both population-differentiating variants as well as long-range LD regions; to capture consecutive outliers that corresponds to long-range LD regions, we apply a Gaussian smoothing to these statistics.

@@

For detecting outlier samples in PCA, we use a modified version of the Probabilistic Local Outlier Factor (PLOF) statistic on PCs \cite[]{kriegel2009loop}. Using K nearest neighbours (KNN), this consists in comparing the distance from an point $j$ to its KNNs ($pd_j = \sqrt{\frac{1}{K} \sum_{k=1}^K d^2_{j \leftrightarrow j_k}}$, where $j_k$ is the $k$-th NN of $j$) with the distances from its KNNs to their KNNs ($\frac{1}{K} \sum_{k=1}^K pd_{j_k}$). The idea is that an outlier should be far from all other points, but this point is even more outlier if its KNNs are in a very dense cluster. 
In practice, we use $pd_j / \sqrt{\frac{1}{K} \sum_{k=1}^K pd_{j_k}}$ (note the square root) as it otherwise detect any point next to a very dense cluster as an outlier.
We implement this statistic in function \texttt{prob\_dist} of R package bigutilsr. To make it fast, we use and parallelise (\texttt{knn\_parallel}) the fast K nearest neighbours implementation of R package nabor \cite[]{elseberg2012comparison}.
Automatic outlier detection is very hard, therefore we recommend users to choose a threshold of outlierness for this statistic based on visual inspection (histogram of these statistics and PC scores colored by these statistics).

For detecting samples that have a different ancestry from most of the samples in the data, we compute the pairwise orthogonalized Gnanadesikan-Kettenrin robust Mahalanobis distances on PC scores (Section \ref{maha}). We then restrict to individuals whose log-distance (alternatively p-value) is smaller (larger) than some threshold determined based on visual inspection.

@@

To project PCs of a reference dataset (e.g.\ 1000G) to a target genotype dataset, we implement the following 3 steps in function \texttt{bed\_projectPCA} of package bigsnpr: 1) matching the variants of each dataset, including removing ambiguous alleles [A/T] and [C/G], and matching strand and direction of the alleles; 2) computing PCA of the reference dataset using the matched variants only; 3) projecting computed PCs to the target data using an optimised implementation (see Supplementary Materials) of the Online Augmentation, Decomposition, and Procrustes (OADP) transformation \cite[]{zhang2019fast}.
To project individuals from the same dataset as the ones used for computing PCA, you can use function \texttt{bed\_projectSelfPCA}. 
Note that the new individuals to be projected should not be related to the ones used for computing PCA.

@@

We provide a subsetted version of the 1000 genomes (1000G) project data \cite[]{10002015global,meyer2019genotype}.
Variants are restricted to the ones in common with HapMap3 or UK Biobank \cite[]{international2010integrating,bycroft2017genome}. 
Moreover, we apply some quality control filters; we remove variants having a minor allele frequency < 0.01, variants with P-value of the Hardy-Weinberg exact test < $10^{-50}$, and non-autosomal variants. 
To remove related individuals with second-degree relationship or more, we apply KING cutoff of 0.0884 to the data using PLINK 2.0 \cite[]{manichaikul2010robust,chang2015second}.
This results in 2490 individuals and 1,664,852 variants of the 1000G project (phase 3) in PLINK bed/bim/fam format. 
Resulting PLINK files and R code to generate these files are available at \url{https://doi.org/10.6084/m9.figshare.9208979.v3}. 
To easily download these data, we provide function \texttt{download\_1000G} in R package bigsnpr.

@@

@@

To show that we provide very fast implementations of the different methods presented in this paper, we apply them to the UK Biobank. We use 20 physical cores for most of the computations.
It takes 22 minutes to perform a first phase of clumping on 406,545 unrelated individuals genotyped over 504,139 variants, which reduces the number of variants to 261,307. It then takes 34 minutes to compute the first 20 PCs using these 261,307 variants.
If performing the automatic procedure for LD detection, it takes 5 hours to perform the initial clumping step, 6 rounds of computation of PCs and 5 rounds of outlier variant detection (i.e.\ 5 iterations of outlier detection and one final computation of PCs).

@@

Finally, it takes only 6 minutes to compute the 30-Nearest Neighbours of 20 PC scores for 406,545 UK Biobank individuals, which is the main computational aspect of computing statistics for detection individual outlier samples (Section \ref{outlier-sample}).

[SHOULD INCLUDE PLINK 2.0 WITH AND WITHOUT AVX2?]

@@

Then, we computed again PCA after restricting the individuals included in the computations: we randomly subsampled UKBB data to use only 10,000 British individuals (out of 431,029) and 5000 Irish individuals (out of 12,755), while keeping all individuals with other self-reported individuals. 
We further removed all pairs of related individuals reported by the UKBB (both individuals in each pair). 
This resulted in 48,942 individuals that we used to compute 50 PCs, which took less than 3 hours using function \texttt{bed\_autoSVD} (that converged after 4 iterations of automatic LD removal). 
We show that we are able to capture at least 40 PCs that display visual population structure (Figures \ref{fig:UKBB-scores2} and \ref{fig:UKBB-screeplot2}).
We then projected all 439,429 remaining individuals from UKBB onto this PCA space in 21 minutes only. Note that these individuals should not be related to any of the 48,942 individuals used for training PCA because we removed both individuals from each pair of related individuals in UKBB. Projection of new individuals show again a clear shrinkage when using simple projection (between 1.00 for PC1 and 1.80 for PC50), but no visible bias when using OADP projection (Figure \ref{fig:projUKBB-related}).

@@

@@

\noindent[ADD DISCU ON OTHER USE + OTHER POSSIBLE PROBLEMS -> METHYL]

@@

\begin{figure}[!htpb]
\centerline{\includegraphics[width=0.8\textwidth]{UKBB-screeplot.pdf}}
\caption{Scree plot: plot of singular values computed on the UK Biobank.
\label{fig:UKBB-screeplot}}
\end{figure}

\begin{figure}[!htpb]
\centerline{\includegraphics[width=0.9\textwidth]{UKBB-PC1-20.png}}
\caption{Principal Component (PC) scores 1 to 20 computed on the UK Biobank.
\label{fig:UKBB-scores}}
\end{figure}

\begin{figure}[!htpb]
\centerline{\includegraphics[width=0.9\textwidth]{UKBB-loadings.pdf}}
\caption{Principal Component (PC) loadings 1 to 20 computed on the UK Biobank.
\label{fig:UKBB-loadings}}
\end{figure}

@@

\begin{figure}[!htpb]
\centerline{\includegraphics[width=0.8\textwidth]{UKBB-screeplot-restricted.pdf}}
\caption{Scree plot: plot of singular values computed on the UK Biobank using 48,942 individuals of diverse ancestries.
\label{fig:UKBB-screeplot2}}
\end{figure}

\begin{figure}[!htpb]
\centerline{\includegraphics[width=0.9\textwidth]{UKBB-scores-restricted.png}}
\caption{Principal Component (PC) scores 27 to 50 computed on the UK Biobank using 48,942 individuals of diverse ancestries. Different colors represent different self-reported ancestries.
\label{fig:UKBB-scores2}}
\end{figure}
